GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2025-06-25 writeUps/paper.tex
Updates from Overleaf

When students get stuck on math problems during homework or classwork, it may be helpful to offer them hints or other ``just-in-time'' support \cite[e.g][]{jit}, possibly helping them overcome specific misconceptions or other sources of confusion and allowing them to move on in the assignment.
The availability of problem-specific support for struggling students is one potential advantage of educational technology over traditional pencil-and-paper practice.
On the other hand, the ready availability of hints may discourage students from working through difficult problems on their own, thereby inhibiting their learning.

@@

Data from the randomized controlled trial (RCT) reported in \citet{impactPaper} may provide evidence for one or another side of this debate.
That RCT contrasted four different applications for computer-based math practice, one of which included bottom-out hints.
If that condition benefited students who frequently requested bottom-out hints (``bottom-outers'') more than those who didn't, that may suggest that the availability of bottom-out hints played a role in the condition's effectiveness.
Alternatively, if students who avoided bottom-out hints benefited more, it may be because the availability of bottom-out hints impeded the condition's effectiveness.
(There are certainly other factors at play as well, so this analysis would not be conclusive.)
Indeed, if the use of bottom-out hints is considered part of the treatment, then those students who rarely used them are partial compliers---could the treatment still benefit them?

@@

\citet{frangakis} introduced the terminology of principal stratification and extended the framework well beyond noncompliance, inspiring a broad literature introducing (and debating) novel applications and estimators.
For instance, \citet{fellerEtAl2016} estimated differential effects of random assignment to the Head Start early childhood care program for children who, if assigned to control, would receive home-based care, and for those who would attend a different early childhood care program.
Other methodologists have developed estimators for the evaluation of surrogate outcomes \citep{li2010bayesian}, the investigation of causal mechanisms \citep{lidsayPage}, sample attrition \citep{zhangRubin, ding2011}, and other problems.
Since interest is often in comparing effects between principal strata---a goal that runs counter to the exclusion restriction---the earliest principal stratification estimators built on \citet{imbens1997bayesian}'s parametric mixture models \citep[e.g.][]{Barnard01062003,mealli2004analyzing}, and that approach remains popular \citep{lidsayPage,fellerEtAl2016}.
However, these models are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
The most common type of parametric model, the normal mixture model, is especially dubious for studying bottom-outers in \citet{impactPaper}'s RCT, whose outcome, the total score on a 10-item posttest, is not normally distributed.

@@

This paper expands on those latter results by presenting a new principal effect estimator for the case of one-way noncompliance, in which principal stratum membership is observed in one treatment group but not the other (future work will extend our approach to general scenarios). Our estimator boils down to a simple, three-step procedure: first, estimate principal scores, for instance, with logistic regression; then, impute unknown principal subgroup membership indicators with principal scores; and finally estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We show that this estimator is the solution to a set of estimating equations based on the combined moments of principal scores, indicators for treatment assignment and observed principal stratum membership in the treatment group, outcomes, and (optionally) covariates, and give a set of assumptions under which it is consistent for the true principal effects.
We will refer to the estimators presented here as ``\textsc{geepers}'': general estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on the correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability and admits to sandwich-style standard error estimates \citep{stefanskiBoos}, which we provide in the appendix.

@@

This work advances the literature in three ways: first, it makes an explicit connection between principal score methods, M-estimation, and semi-parametric estimation of principal effects using auxiliary data. Second, it provides a general estimator that relies on two of the most familiar tools in the applied education researcher's toolbox: linear and logistic regression. The \textsc{geepers} method can, therefore, make principal stratification methods accessible for a much broader range of applied education researchers. Lastly, this work includes an extensive simulation study that compares \textsc{geepers} to parametric mixture modeling and principal score weighting under a range of favorable and challenging scenarios. The simulation demonstrates the strengths and weaknesses of each method and the tradeoffs between them.

@@

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken after treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exhibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance''):%\citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff0=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is because strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

This approach has several problems.
First, the assumption of a conditional parametric model %normality assumption 
is untestable and limiting.
Unfortunately, even when it holds, mixture model estimates can be unreliable and, in some cases, severely biased \citep{griffin2008application,feller2017principal}.

@@

Principal scores can potentially improve inference in a finite mixture model, by re-writing \eqref{eq:mixtureModel} as
\begin{equation}\label{eq:mixtureModelPS}
  f_{Y_C}(y|\bm{x})=(1-\ppip{\bm{\alpha}})%\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)
g^0(y|\bm{x})+\ppip{\bm{\alpha}}%\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
g^1(y|\bm{x})
\end{equation}
where $g^s(y|\bm{x})$, $s=0,1$, are conditional densities which must be specified. Most commonly, researchers assume a linear normal model, e.g., $g^s(y|\bm{x})=\phi\left(\frac{y-\mucs-\bm{\gamma}'\bm{x}}{\sigma_C^s}\right)$, where the columns of $\bm{x}$ are centered so that $\mathbb{E}[\bm{x}]=\bm{0}$. 
%Ideally, a statistician will estimate parameters $\muc0$, $\muc1$, $\mut0$, $\mut1$, $\sigma^0_C$, $\sigma^1_C$, and $\bm{\alpha}$ simultaneously, using maximum likelihood or Bayesian techniques.

@@

In each run of the simulation, we simulated three independent standard normal covariates, $x_k$ $k=1,\dots,3$. However, only the first two covariates were ``observed,'' i.e., were included in the analysis model.
Given all three covariates, the principal scores were set as
\begin{equation*}
  \ppi=logit^{-1}\left[\alpha(x_{1i}-x_{2i}+x_{3i})\right]
\end{equation*}
where $\alpha$ is a manipulated factor.
When $\alpha$ was higher, $\st$ was more easily predicted by covariates; when $\alpha=0$, Assumption \ref{ass:vps} was violated, i.e., $Var(\pp)=0$.
Principal stratum $\sti\in \{0,1\}$ was simulated as $S_i\sim Bern(\ppi)$.

@@

The methods all used the same ``observed'' dataset, consisting of outcomes $Y$, treatment assignments $Z$, two covariates $\bxs=\bxy=\bx=[x_1,x_2]$, and $S$ observed only when $Z=1$.
The third covariate used in the data generating model, $x_3$, was ``unobserved''--since $x_3$ was correlated with both $\st$ and $Y$, Principal Ignorability \eqref{eq:pi} was violated.

@@

We omitted the $\alpha=0$ case from Figures \ref{fig:alphan} and \ref{fig:boxplots} since the variance of \geepers and \pmm estimates was so large as to dominate the plot. 
However, the first column of results in Table \ref{tab:coverage} shows that, when other assumptions are met, CIs from both methods tend to over-cover when $\alpha=0$. 
In a randomized experiment, $\alpha$ can be estimated directly, making Assumption \ref{ass:vps} testable.

@@

\label{evalMetric}The simulation study suggests that none of the three methods evaluated---\geepers, \pmm, or \psw---dominates the others, and offers some guidance on when each estimator may be preferable. \psw is the most precise and requires no assumptions about the distribution of outcomes, but can be quite sensitive to violations of principal ignorability. When the determinants of $\st$ are well-understood and measured, \psw seems to be the best option. When principal ignorability does not hold, potential outcomes are conditionally normal, and there are no $X:Z$ interactions within principal strata, the \pmm estimator outperforms \geepers, though the difference decreases with $n$ or $\alpha$. \geepers may be the best choice when conditional normality may not hold, or if there are $X:Z$ interactions and $\bm{x}$ is sufficiently predictive of $\st$ (AUC $\ge$0.675, corresponding to $\alpha \ge 0.5$, may serve as a threshold). 
These suggestions may depend heavily on the simulation setup here; applying them in general scenarios requires more research.

@@

\section{Applications}
To illustrate \textsc{geepers} and compare it to the other discussed methods, we include two applied examples.
First, we briefly replicate an analysis from \citet{richardson2023estimating} using the \textsc{geepers} and the other methods discussed above.
This is a relatively straightforward example of an experiment with imperfect compliance in which \citet{richardson2023estimating} estimated statistically significant effects overall in both principal strata.
Next, we present a novel analysis of data from the RCT of computer-based math applications reported in \citet{impactPaper}, and look into the role of bottom-out hints, as described in the introduction.
This analysis aims to shed light on an important question in education research, illustrates model and covariate selection, and includes a discussion on the plausibility of identification assumptions.

@@

The results are displayed in Figure \ref{fig:opt}% and Table \ref{tab:optEffs}
: both \textsc{geepers} and the Bayesian mixture model gave similar results to those reported in \citet{richardson2023estimating} from BSIV.
For the principal stratum of subjects who would not fully complete the treatment, those three methods estimated that assignment to the treatment condition reduced percentage of dental sites with bleeding by 31--34 percentage points; for subjects who would complete the treatment, the three estimated a reduction of 14-16 percentage points.
The conclusion of these methods is somewhat surprising: that the effect was larger for subjects who would not complete the treatment than for those who would (the two strata had similar bleeding percentages at baseline, and a logit transformation of the outcome leaves the results qualitatively unchanged). 
One possible explanation is the conjecture that subjects stop showing up for treatment once their periodontal disease has abated.

@@

\subsubsection{Principal Stratification and Alternatives}\label{sec:psInEdTech}
The goal of principal strafication in this context is to determine if assignment to the ASSISTments condition affected students who tend to request frequent bottom-out hints differently than students who tended to request fewer bottom-out hints.
Is this the right question to ask?

@@

Another arguable disadvantage of the observational study approach is that it overlooks the broader context of the experiment, focusing on whether bottom-out hints affect posttest scores rather than examining their role in ASSISTments' overall effectiveness.
A second approach, mediation analysis, contextualizes the bottom-out effect within the larger experiment.
It, too, relies on the same non-confounding assumption as an observational study, and also suffers from an additional challenge: mediation analysis relies on a ``common support'' condition that ``the mediator state must not be a deterministic function of the treatment'' \citep[p. 225]{celli2022causal}.
However, bottom-out hints are unavailable in the other three arms of the study, violating this condition (\citealt{sales2021student} offers a solution, which we find unconvincing).

@@

Lastly, Assumption \ref{ass:ci} implies that there is no interaction between $\bxs$ and $\st$ in the outcome model. Further, for the sake of statistical efficiency, we did not consider outcome models with interactions between covariates and $Z$.
\emph{A priori}, the student attribute that poses the greatest threat to these modeling choices is prior achievement, that is, scores on the pretest and the fifth-grade state test.
This is both because they measure a similar construct to the posttest, and because \cite{impactPaper} found a pretest-treatment interaction in a different treatment contrast.
Following that reasoning, we conducted a third parallel analysis excluding prior achievement covariates.% in $\bxs$ but including all covariates in $\bxy$.

@@

Outcome models for all analyses adjusted for all of the variables that were in the corresponding principal score models, in addition to measurements of students' prior knowledge; also, the models substituted class fixed effects for school fixed effects.\footnote{Why use a coarser set of fixed effects for the principal score model than the outcome regression? First, the principal score model was fit to a smaller sample (only students in the ``Instant'' condition), so some classrooms only included one student, which can lead to poor fit. Second, the asymptotics of logistic regression are more restricted than OLS, so logistic regression tends to perform poorly when the number of parameters increases with the sample size \citep{agresti}.}
As with principal score models, we checked model fit with residual plots and modified model specifications accordingly.

@@

The Bayesian mixture model assumed that outcomes were normally distributed, conditional on covariates and principal stratum, with a residual standard deviation that varied between principal strata. This modeling assumption is necessarily false since post-test scores were integers between 0 and 10. All model parameters with support in $\mathbb{R}$ were given standard normal priors, while standard deviation parameters were given half-standard normal priors.
The principal score and outcome models were fit simultaneously using a Markov Chain Monte Carlo algorithm \citep{rstan}.

@@

The other estimation strategies largely agreed with \textsc{geepers} and with each other in the contrast between ASSISTments and BAU, and, to a lesser extent, in the contrast between ASSISTments and DragonBox. The biggest difference between the methods was in the contrast with FH2T, in which the Bayesian mixture model resulted in a larger difference between the two principal effects, as well as the largest standard errors, while the two \textsc{psw} principal effect estimates were quite similar.  However, these differences are difficult to interpret due to the large uncertainty around all of the point estimates.

@@

There is good reason to hope that extensions to \textsc{geepers}, including cases in which $S$ takes more than two values, may be straightforward.
For instance, in truncation by death problems \citep[e.g.][]{zhangRubin,ding2011} with weak monotonicity, $S_i=1$ if participant $i$ survives (or, more generally, if the outcome $Y_i$ is measured) and 0 otherwise, interest is typically in the principal effect for the ``always survive'' principal stratum in which $\sti=S_{ci}=1$.
If, say, $\sti\ge S_{Ci}$, then every subject in the control condition with $S_i=1$ is in the always survive stratum, while those subjects in the treatment condition with $S_i=1$ are a mixture of the always survive and $\sti=1;\;S_{Ci}=0$ stratum.
This scenario is broadly similar to the strong monotonicity, one-way noncompliance situation that this paper discussed.
On the other hand, when no monotonicity assumption holds, both $\st$ and $S_C$ will have to be imputed for every subject; further research is necessary to determine the appropriate way to do so.

@@

Sampling error prevents firm conclusions regarding principal effects for students who would, if given the opportunity, request frequent bottom-out hints.
However, the results are consistent with the hypothesis that ASSISTments is less effective for frequent bottom-out requesters when compared to a similar program without hints or immediate feedback. We hypothesize that many students who request bottom-out hints frequently are gaming the system, abetted by the availability of bottom-out hints.

>>>>

When students get stuck on math problems during homework or classwork, it may be helpful to offer them hints or other ``just-in-time'' support \cite[e.g.,][]{jit}, possibly helping them overcome specific misconceptions or other sources of confusion and allowing them to move on in the assignment.
The availability of problem-specific support for struggling students is one potential advantage of educational technology over traditional pencil-and-paper practice.
On the other hand, the ready availability of hints may discourage students from working through difficult problems independently, thereby hindering their learning.

@@

Data from the randomized controlled trial (RCT) reported in \citet{impactPaper} may provide evidence for one or another side of this debate.
That RCT contrasted four different applications for computer-based math practice, one of which included bottom-out hints.
If that condition benefited students who frequently requested bottom-out hints (``bottom-outers'') more than those who didn't, that may suggest that the availability of bottom-out hints played a role in the condition's effectiveness.
Alternatively, if students who avoided bottom-out hints benefited more, it may be because the availability of bottom-out hints impeded the condition's effectiveness.
(There are certainly other factors at play as well, so this analysis would not be conclusive.)
Indeed, if the use of bottom-out hints is considered part of the treatment, then those students who rarely used them were partial compliers---did the treatment still benefit them?

@@

\citet{frangakis} introduced the terminology of principal stratification and extended the framework well beyond noncompliance, inspiring a broad literature introducing (and debating) novel applications and estimators.
For instance, \citet{fellerEtAl2016} estimated differential effects of random assignment to the Head Start early childhood care program for children who, if assigned to control, would receive home-based care, and for those who would attend a different early childhood care program.
Other methodologists have developed estimators for evaluating surrogate outcomes \citep{li2010bayesian}, investigating causal mechanisms \citep{lidsayPage}, addressing sample attrition \citep{zhangRubin, ding2011}, and other related problems.
Since interest is often in comparing effects between principal strata---a goal that runs counter to the exclusion restriction---the earliest principal stratification estimators built on \citet{imbens1997bayesian}'s parametric mixture models \citep[e.g.][]{Barnard01062003,mealli2004analyzing}, and that approach remains popular \citep{lidsayPage,fellerEtAl2016}.
However, these models are often difficult to specify and can be either hard or impossible to test.
Even when they are well-specified, they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
The most common type of parametric model, the normal mixture model, is especially dubious for studying bottom-outers in \citet{impactPaper}'s RCT, whose outcome, the total score on a 10-item posttest, is not normally distributed.

@@

This paper expands on those latter results by presenting a new principal effect estimator for the case of one-way noncompliance, in which principal stratum membership is observed in one treatment group but not the other (future work will extend our approach to general scenarios). Our estimator boils down to a simple, three-step procedure: first, estimate principal scores, for instance, with logistic regression; then, impute unknown principal subgroup membership indicators with principal scores; and finally, estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We show that this estimator is the solution to a set of estimating equations based on the combined moments of principal scores, indicators for treatment assignment and observed principal stratum membership in the treatment group, outcomes, and (optionally) covariates, and give a set of assumptions under which it is consistent for the true principal effects.
We will refer to the estimators presented here as ``\textsc{geepers}'': general estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on the correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability and admits to sandwich-style standard error estimates \citep{stefanskiBoos}, which we provide in the appendix.

@@

This work advances the literature in three ways: first, it makes an explicit connection between principal score methods, M-estimation, and semi-parametric estimation of principal effects using auxiliary data. Second, it provides a general estimator that relies on two of the most familiar tools in the applied education researcher's toolbox: linear and logistic regression. The \textsc{geepers} method can, therefore, make principal stratification methods accessible for a much broader range of applied education researchers. Lastly, this work includes an extensive simulation study that compares \textsc{geepers} to parametric mixture modeling and principal score weighting under a range of favorable and challenging scenarios. The simulation demonstrates the strengths and weaknesses of each method, as well as the trade-offs between them.

@@

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken after treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exhibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance''):%\citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff0=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

This approach has several problems.
First, the assumption of a conditional parametric model is untestable and limiting.
Unfortunately, even when it holds, mixture model estimates can be unreliable and, in some cases, severely biased \citep{griffin2008application,feller2017principal}.

@@

Principal scores can potentially improve inference in a finite mixture model by modifying \eqref{eq:mixtureModel} as
\begin{equation}\label{eq:mixtureModelPS}
  f_{Y_C}(y|\bm{x})=(1-\ppip{\bm{\alpha}})%\phi\left(\frac{y-\muc0}{\sigma_C^0}\right)
g^0(y|\bm{x})+\ppip{\bm{\alpha}}%\phi\left(\frac{y-\muc1}{\sigma_C^1}\right)
g^1(y|\bm{x})
\end{equation}
where $g^s(y|\bm{x})$, $s=0,1$, are conditional densities which must be specified. Most commonly, researchers assume a linear normal model, e.g., $g^s(y|\bm{x})=\phi\left(\frac{y-\mucs-\bm{\gamma}'\bm{x}}{\sigma_C^s}\right)$, where the columns of $\bm{x}$ are centered so that $\mathbb{E}[\bm{x}]=\bm{0}$. 
%Ideally, a statistician will estimate parameters $\muc0$, $\muc1$, $\mut0$, $\mut1$, $\sigma^0_C$, $\sigma^1_C$, and $\bm{\alpha}$ simultaneously, using maximum likelihood or Bayesian techniques.

@@

In each run of the simulation, we simulated three independent standard normal covariates, $x_k$, $k=1,\dots,3$. However, only the first two covariates were ``observed,'' i.e., were included in the analysis model.
Given all three covariates, the principal scores were set as
\begin{equation*}
  \ppi=logit^{-1}\left[\alpha(x_{1i}-x_{2i}+x_{3i})\right]
\end{equation*}
where $\alpha$ is a manipulated factor.
When $\alpha$ was higher, $\st$ was more easily predicted by covariates; when $\alpha=0$, Assumption \ref{ass:vps} was violated, i.e., $Var(\pp)=0$.
Principal stratum $\sti\in \{0,1\}$ was simulated as $S_i\sim Bern(\ppi)$.

@@

The methods all used the same ``observed'' dataset, consisting of outcomes $Y$, treatment assignments $Z$, two covariates $\bxs=\bxy=\bx=[x_1,x_2]$, and $S$ observed only when $Z=1$.
The third covariate used in the data-generating model, $x_3$, was ``unobserved''--since $x_3$ was correlated with both $\st$ and $Y$, Principal Ignorability \eqref{eq:pi} was violated.

@@

We omitted the $\alpha=0$ case from Figures \ref{fig:alphan} and \ref{fig:boxplots} since the variance of \geepers and \pmm estimates was so large as to dominate the plot. 
However, the first column of results in Table \ref{tab:coverage} shows that, when other assumptions are met, confidence intervals from both methods tend to over-cover when $\alpha=0$. 
In a randomized experiment, $\alpha$ can be estimated directly, making Assumption \ref{ass:vps} testable.

@@

\label{evalMetric}The simulation study suggests that none of the three methods evaluated---\geepers, \pmm, or \psw---dominates the others, and offers some guidance on when each estimator may be preferable. \psw is the most precise and requires no assumptions about the distribution of outcomes, but can be quite sensitive to violations of principal ignorability. When the determinants of $\st$ are well-understood and measured, \psw seems to be the best option. When principal ignorability does not hold, potential outcomes are conditionally normal, and there are no $X:Z$ interactions within principal strata, the \pmm estimator outperforms \geepers, though the difference decreases with $n$ or $\alpha$. \geepers may be the best choice when conditional normality may not hold, or if there are $X:Z$ interactions and $\bm{x}$ is sufficiently predictive of $\st$ (AUC $\ge$0.675, corresponding to $\alpha \ge 0.5$, may serve as a threshold). 
These suggestions may depend heavily on the simulation setup; applying them in general scenarios requires further research.

@@

\section{Applications}
To illustrate \textsc{geepers} and compare it to the other discussed methods, we include two applied examples.
First, we briefly replicate an analysis from \citet{richardson2023estimating} using the \textsc{geepers} and the other methods discussed above.
This is a relatively straightforward example of an experiment with imperfect compliance in which \citet{richardson2023estimating} estimated statistically significant effects overall in both principal strata.
Next, we present a novel analysis of data from the RCT of computer-based math applications reported in \citet{impactPaper}, and look into the role of bottom-out hints, as described in the introduction.
This analysis aims to shed light on an important question in education research, illustrate model and covariate selection, and include a discussion on the plausibility of identification assumptions.

@@

The results are displayed in Figure \ref{fig:opt}% and Table \ref{tab:optEffs}
: both \textsc{geepers} and the Bayesian mixture model gave similar results to those reported in \citet{richardson2023estimating} from BSIV.
For the principal stratum of subjects who would not fully complete the treatment, those three methods estimated that assignment to the treatment condition reduced percentage of dental sites with bleeding by 31--34 percentage points; for subjects who would complete the treatment, the three estimated a reduction of 14-16 percentage points.
The conclusion of these methods is somewhat surprising: that the effect was larger for subjects who would not complete the treatment than for those who would (the two strata had similar bleeding percentages at baseline, and a logit transformation of the outcome leaves the results qualitatively unchanged). 
One possible explanation is the conjecture that subjects stopped attending treatment once their periodontal disease improved.

@@

\subsubsection{Principal Stratification and Alternatives}\label{sec:psInEdTech}
The goal of principal stratification in this context is to determine if assignment to the ASSISTments condition affected students who tended to request frequent bottom-out hints differently than students who tended to request fewer bottom-out hints.
Is this the right question to ask?

@@

Another arguable disadvantage of the observational study approach is that it overlooks the broader context of the experiment, focusing solely on whether bottom-out hints affect posttest scores rather than examining their role in the overall effectiveness of ASSISTments.
A second approach, mediation analysis, contextualizes the bottom-out effect within the larger experiment.
It, too, relies on the same non-confounding assumption as an observational study, and also suffers from an additional challenge: mediation analysis relies on a ``common support'' condition that ``the mediator state must not be a deterministic function of the treatment'' \citep[p. 225]{celli2022causal}.
However, bottom-out hints are unavailable in the other three arms of the study, violating this condition (\citealt{sales2021student} offers a solution, which we find unconvincing).

@@

Lastly, Assumption \ref{ass:ci} implies that there is no interaction between $\bxs$ and $\st$ in the outcome model. Further, for the sake of statistical efficiency, we did not consider outcome models with interactions between covariates and $Z$.
\emph{A priori}, the student attribute that poses the greatest threat to these modeling choices is prior achievement, that is, scores on the pretest and the fifth-grade state test.
This is both because they measure a similar construct to the posttest and because \cite{impactPaper} found a pretest-treatment interaction in a different treatment contrast.
Following that reasoning, we conducted a third parallel analysis excluding prior achievement covariates. % in $\bxs$ but including all covariates in $\bxy$.

@@

Outcome models for all analyses included for all of the variables that were in the corresponding principal score models, in addition to measurements of students' prior knowledge; also, the models substituted class fixed effects for school fixed effects.\footnote{Why use a coarser set of fixed effects for the principal score model than the outcome regression? First, the principal score model was fit to a smaller sample (only students in the ``Instant'' condition), so some classrooms only included one student, which can lead to poor fit. Second, the asymptotics of logistic regression are more restricted than OLS, so logistic regression tends to perform poorly when the number of parameters increases with the sample size \citep{agresti}.}
As with principal score models, we checked model fit with residual plots and modified model specifications accordingly.

@@

The Bayesian mixture model assumed that outcomes were normally distributed, conditional on covariates and principal strata, with a residual standard deviation that varied between principal strata. This modeling assumption is necessarily false since post-test scores were integers between 0 and 10. All model parameters with support in $\mathbb{R}$ were given standard normal priors, while standard deviation parameters were given half-standard normal priors.
The principal score and outcome models were fit simultaneously using a Markov Chain Monte Carlo algorithm \citep{rstan}.

@@

The other estimation strategies largely agreed with \textsc{geepers} and with each other in the contrast between ASSISTments and BAU, and, to a lesser extent, in the contrast between ASSISTments and DragonBox. The biggest difference between the methods was in the contrast with FH2T, in which the Bayesian mixture model resulted in a larger difference between the two principal effects, as well as the largest standard errors, while the two \textsc{psw} principal effect estimates were quite similar.  However, these differences are difficult to interpret due to the large uncertainty surrounding all the point estimates.

@@

There is good reason to hope that extensions to \textsc{geepers}, including cases in which $S$ takes more than two values, may be straightforward.
For instance, in truncation by death problems \citep[e.g.][]{zhangRubin,ding2011} with weak monotonicity, $S_i=1$ if participant $i$ survives (or, more generally, if the outcome $Y_i$ is measured) and 0 otherwise, interest is typically in the principal effect for the ``always survive'' principal stratum in which $\sti=S_{ci}=1$.
If, say, $\sti\ge S_{Ci}$, then every subject in the control condition with $S_i=1$ is in the always survive stratum, while those subjects in the treatment condition with $S_i=1$ are a mixture of the always survive and $\sti=1;\;S_{Ci}=0$ stratum.
This scenario is broadly similar to the strong monotonicity, one-way noncompliance situation discussed in this paper.
On the other hand, when no monotonicity assumption holds, both $\st$ and $S_C$ will have to be imputed for every subject; further research is necessary to determine the appropriate way to do so.

@@

Sampling error prevents firm conclusions regarding principal effects for students who would, if given the opportunity, request frequent bottom-out hints.
However, the results are consistent with the hypothesis that ASSISTments is less effective for frequent bottom-out requesters when compared to a similar program without hints or immediate feedback. We hypothesize that many students who frequently request bottom-out hints are gaming the system, facilitated by the availability of these hints.
